---
title: "Reproducible Research: Peer Assessment 1"
author: "Philo Bishay"
date: "December 20, 2015"
output: html_document
keep_md: true
---
# Knitr with R Markdown
## Loading and preprocessing the data

```{r load_data, echo = TRUE}

#### Optional Code -- specify work directory ###########
# setwd("Y:/Documents/Coursera/Reproducible Research/Assignment")
########################################################

download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              dest ="data.zip")

#data will be a dataframe containing the raw data
data = read.delim(unzip("data.zip"),sep = ",",na.strings = "NA")

#data2 will be a dataframe with processed data
data2 = data
data2$steps = as.integer(data$steps)
data2$date = strptime(data$date,"%Y-%m-%d")
str(data2)

```


## What is mean total number of steps taken per day?
```{r steps_per_day, echo = TRUE}

hist(data2$steps,main = "Total steps taken each day",
     ylab="Frequency",xlab="Steps taken", col="red")
#mean and median:
  step.summary = summary(data2$steps)
  step.summary[c(4,3)]
```
### Note, there is a large point mass at zero, so I show the histogram a second time excluding zero-step data.

```{r second_histogram, echo = TRUE}
  hist(data2$steps[data2$steps>0],main = "Total steps excluding zeroes", ylab="Frequency",xlab="Steps taken", col="green")
  
```

## What is the average daily activity pattern?
```{r avg_daily_activity, echo = TRUE}
steps.by.interval = aggregate(steps~interval,data=data2,FUN=mean)
names(steps.by.interval)[2] = "Avg.steps"

plot(Avg.steps~interval,data=steps.by.interval,type="l",main="Average Steps each interval", ylab="average steps")

# Print the max number of steps
max.steps = max(steps.by.interval$Avg.steps)
max.interval = steps.by.interval[steps.by.interval$Avg.steps==max.steps,]
cat("Interval #", max.interval$interval , " contains " , max.interval$Avg.steps , " steps, the maximum.",sep="")
```


## Imputing missing values
### First, we count the NA's:

```{r missing_vals, echo = TRUE}
cat("There are",sum(is.na(data2)), "missing values")
```

### Next, use the 5-min interval mean to fill in missing values:

```{r fill_in_vals, echo = TRUE}
#fill "NA" step values with interval averages:
filled.steps = data2$steps
na.intervals = data.frame(data2[is.na(filled.steps),3])
names(na.intervals) = "interval"
fill.data = merge(na.intervals,steps.by.interval)
filled.steps = replace(filled.steps,is.na(filled.steps),fill.data$Avg.steps)
summary(filled.steps)
```

### Create new dataset filling in missing data:
```{r new_data, echo = TRUE}
data2 = cbind(data2,filled.steps)
summary(data2)
cat("The means of both datasets should be the same. Old:", mean(data2$steps,na.rm=T),"New:",mean(data2$filled.steps))
```

### Compare to original histogram
```{r comparison_to_first_histogram, echo = TRUE}
data2$factor.date = data$date

hist(data2$filled.steps,main = "Total steps taken each day",
     ylab="Frequency",xlab="Steps taken", col="blue")


#mean and median:
  step.summary2 = summary(data2$filled.steps)
  step.summary2[c(4,3)]
  cat("For comparison, the difference between the original means and medians and those of the revised data set is",step.summary2[c(4,3)] - step.summary[c(4,3)],", respectively. This is because NA's were replaced with means of the original data.")
```

## Are there differences in activity patterns between weekdays and weekends?
### First, differentiate weekends from weekdays.
```{r weekdays, echo = TRUE}
data2$day = weekdays(data2$date)
data2$weekday.or.weekend = data2$day
is.weekend = data2$weekday.or.weekend %in% c("Sunday","Saturday")
data2$weekday.or.weekend = replace(data2$weekday.or.weekend,is.weekend,"Weekend")
data2$weekday.or.weekend = replace(data2$weekday.or.weekend,!is.weekend,"Weekday")
data2$weekday.or.weekend = as.factor(data2$weekday.or.weekend)
cat("Number of weekdays and weekends respectively:",summary(data2$weekday.or.weekend))
```
### Next plot steps by weekend/ weekday
```{r plot_weekends, echo=TRUE}

require(lattice)
summary.data = aggregate(filled.steps~interval+weekday.or.weekend, data=data2, FUN=mean)
xyplot(filled.steps ~ interval | weekday.or.weekend, data = summary.data, layout = c(1, 2), type = "l",
       xlab="5-minute Intervals", ylab="Number of Steps",
       main="Steps on Weekends and Weekdays")
```

```{r knit2HTML, echo=FALSE}
knit2html("PA1_template.Rmd")
```